
---
title: "Microbiome pipeline QC template"  
author: "Your name"  
date: "Month date, year"  
output: html_document  
---

**Download this file and open in RStudio**

# Microbiome QC and prelimanary analysis Pipeline  

In all cases, the first thing that needs to be done is to check the quality of your data. For this, the **microbiomeutilities** package provides a single function called `run_microbiome_pipeline` that puts together `base`, `phyloseq` and `microbiome` R package functions to provide the user with data qc visualisation. In addition, this function saves outputs in following folder:  

1. **QC**  
2. **AlphaDiversity**  
3. **Ordinations**  
4. **Others**  
5. **PhyloseqObjects**  

load the required libraries
```{r}
library(RColorBrewer)
library(ape)
library("microbiomeutilities")
library(data.table)

```

Once you have these then run the following command.  
```{r, warning=TRUE, message=TRUE}


run_microbiome_pipeline(biom = "myBiom.biom",
                        mapping = "myMapping.csv",
                        taxonomy = NULL,
                        treefilename = "myTree.tre",
                        type = "biom",
                        out_dir = "F:/PATH/to/my/output/directory",
                        VariableA = "MY_MainVariable",
                        VariableB = "MY_SecondaryVariable",
                        Distmethod = "bray",
                        OrdinatMethod = "MDS",
                        QC = TRUE,
                        filterCount = 4,
                        filterPrev = 0.01,
                        transformation = FALSE,
                        col.palette = "Set2",
                        filterpseq = TRUE, samsize=0)

```

After filling in all the required parameters in th functions you can click on **kinit** on the top of your R studio console.  

The use of Rmarkdown is that once you knit it to a desired file type, html, pdf, doc etc. you can see all the QC messages.   

After successfully running this you can check the folders for output and then make appropriate choice for in-depth analysis.  

**Credits to developers of**:  
Phyloseq  
microbiome  
data.table  
ape  
vegan  
ggplot2  
ggpubr  

For issues please comment on the [microbiomeutilities issues page](https://github.com/microsud/microbiomeutilities/issues)  




