---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to syncomR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```


<font color="red">microbiomeutilities R package is under active development! Store a local copy of the package (download from GitHub) as backup</font>

The microbiomeutilities R package is mainly a wrapper tool with diverse functions for data handling and visualisation. This package has a "`pipeline`" function called `microbiome_pipeline` for carrying out preliminary QC, Alpha Diversity, Ordination and Composition analysis of OTU tables. The output is a HTML report for convenient first-hand investigating of the microbial profilling datadata.  

## Install  

```{r, eval=FALSE}

install.packages("devtools")
devtools::install_github("microsud/microbiomeutilities")
```  


```{r, message=FALSE, warning=FALSE}

library(microbiomeutilities)
library(microbiome)
library(knitr)
library(tibble)
```


## Example data 

Package data from [Zackular et al., 2014: The Gut Microbiome Modulates Colon Tumorigenesis](http://mbio.asm.org/content/4/6/e00692-13).  

Useful resources:  

For more information on phyloseq data structure and uses you can have a look at [Phyloseq](http://joey711.github.io/phyloseq/)  

Tools for microbiome analysis in R. **Microbiome package** URL: [microbiome package](http://microbiome.github.com/microbiome).  


```{r}

data("zackular2014")
ps0 <- zackular2014

print(ps0)
```


## Microbiome analysis pipeline  

Function `microbiome_pipeline` generates an `HTML` report with preliminary QC, Alpha Diversity, Ordination and Composition analysis of OTU tables. This function saves all intermediate files incuding figures and phyloseq objects in user specified directory.

```{r, eval=FALSE}

microbiome_pipeline(
  otufile = "my.biom",
  mapping = "mymap.csv",
  taxonomy = NULL,
  treefilename = "myTree.tre",
  type = "biom",
  work_dir = "F:/path/my/input/filefolder",
  out_dir = "F:/path/to/save/my/files/folder",
  VariableA = "MC_type1",
  VariableB = "Region",
  UnConstOrd = TRUE,
  heatmap = TRUE,
  filterCount = 4,
  filterPrev = 0.01,
  col.palette = "Paired",
  filterpseq = TRUE,
  samsize = NA,
  projectname = "Mock",
  author = "Sudarshan"
)
```

## Formatting the Phyloseq Object  

Most commonly it is observed that the taxonomy file has classification until a given taxonomic level. We will fill the empty cells with the maximum classification available along with the OTU number.  

Check the taxonomy in phyloseq object.  

```{r}

kable(head(tax_table(ps0)))
```

Some have only g__ of s__ information.  

```{r, eval=FALSE}

data("zackular2014")
p0 <- zackular2014
# reduce size for example
ps0 <- core(ps0, detection = 10, prevalence = 20 / 100)
p0.f <- format_phyloseq(p0)
# head(tax_table(p0.f))

kable(head(tax_table(ps0.f))[3:6])
```

Now the available taxonomy is added.  
There is a second version which will change the names in both otu table and taxonomy table. This can be useful if the analysis has to be done at OTU level. Only ID are less useful.    


```{r}
# reduce size for example
ps0 <- core(ps0, detection = 10, prevalence = 20 / 100)

ps0.f2 <- format_to_besthit(ps0)

kable(head(tax_table(ps0.f2))[3:6])
```

As can be seen, the rownames have the OTUIDs and available toxonomic name(s).  

## Summarize the percent taxa classification for phyloseq  

This can be useful to get an overview of taxonomic classifications. Only patterns such as [g__] or `NA` is expected. [g__<empty>] or similar are not considered. Pease convert for eg. g__unclassified to uniform [g__] or simply `NA`.  

```{r eval=FALSE}

percent_classified(ps0)

```


## Distribution of reads  

Useful for QC purposes. Check for siatribution of sequencing depth.     

```{r, eval=FALSE}

p <- plot_read_distribution(ps0, groups = "DiseaseState", 
                            plot.type = "density")
print(p)
```


## Convert phyloseq object to long data format  

Useful if the user wants to plot specific features.  

```{r}

# reduce size for example
ps0 <- core(ps0, detection = 10, prevalence = 20 / 100)

pseq_df <- phy_to_ldf(ps0, transform.counts = NULL)

kable(head(pseq_df))
```

## Plot alpha diversities  

Utility plot function for diversity measures calculated by `microbiome` package.  

```{r, fig.width=8, fig.height=6, message=FALSE}

library(microbiome)
data("zackular2014")
ps0 <- zackular2014

p <- plot_alpha_diversities(ps0,
                            type = "evenness",
                            index.val = "all",
                            plot.type = "stripchart",
                            variableA = "DiseaseState",
                            palette = "Paired"
)

print(p)
```


## Plot ordination and core  

```{r, fig.height=8, fig.width=12, eval=FALSE}

library(microbiomeutilities)
library(RColorBrewer)
data("zackular2014")
p0 <- zackular2014


ps1 <- format_to_besthit(p0)

ps1 <- subset_samples(ps1, DiseaseState == "H")
ps1 <- prune_taxa(taxa_sums(ps1) > 0, ps1)
prev.thres <- seq(.05, 1, .05)
det.thres <- 10^seq(log10(1e-4), log10(.2), length = 10)
pseq.rel <- microbiome::transform(ps1, "compositional")
# reduce size for example
pseq.rel <- core(pseq.rel, detection = 0.001, prevalence = 20 / 100)

ord.bray <- ordinate(pseq.rel, "NMDS", "bray")

p <- plot_ordiplot_core(pseq.rel, ord.bray,
                        prev.thres, det.thres,
                        min.prevalence = 0.9,
                        color.opt = "DiseaseState", shape = NULL, Sample = TRUE
)

print(p)
```

## Plot taxa boxplot  

Plot relative abundance of top taxa specified by user. 

```{r,fig.width=6, fig.height=4, eval=FALSE}

pn <- plot_taxa_boxplot(ps0,
                        taxonomic.level = "Phylum",
                        top.otu = 5, VariableA = "DiseaseState",
                        title = "Relative abudance plot", color = "Set2"
)

print(pn)
```

## Plot taxa barplot  

```{r, fig.width=12, fig.height=4}

library(RColorBrewer)
library(dplyr)
library(microbiomeutilities)

# reduce data for egample purposes
ps.cor <- core(ps0, detection = 10, prevalence = 0.5)

p <- plot_taxa_composition(ps.cor, taxonomic.level = "Phylum", 
                           transform = "compositional", average_by = NULL,
                           palette = brewer.pal(12, "Set2"))
print(p)
```

## Heatmap using phyloseq and pheatmap   

Useful for visualisng differences in top otus between sample groups.  

```{r, fig.width=10, eval=FALSE}
library(pheatmap)
heat.sample <- plot_taxa_heatmap(ps0,
                                 subset.top = 20,
                                 VariableA = "DiseaseState",
                                 heatcolors = rev(brewer.pal(6, "Spectral")),
                                 transformation = "compositional",
                                 cluster_cols = F)
```

## MicrobiomeHD datasets as phyloseq objects  

We provide access to a subset of studies included in the `MicrobiomeHD` database from Duvallet et al 2017: [Meta-analysis of gut microbiome studies identifies disease-specific and shared responses](https://www.nature.com/articles/s41467-017-01973-8#ref-CR33). Nature communications.    

The phyloseq objects are stored and accessed from [microbiomedatarepo](https://github.com/microsud/microbiomedatarepo).  

```{r}

study <- list_microbiome_data(printtab = FALSE)

knitr::kable(study)
```


Below is the per study reference.  

**NOTE**: When using these studies, please cite [Duvallet et al. 2017](https://www.nature.com/articles/s41467-017-01973-8#ref-CR33) and the respective studies.  

```{r, eval=FALSE}

file <- system.file("extdata", "microbiomeHD_ref.txt", package = "microbiomeutilities")
reference <- read.table(file, header = T, sep = "\t")

knitr::kable(reference)
```

For more tutorials and examples of data anlaysis in R please check:  

* [Microbiome data analysis SpringSchool2018](https://goo.gl/CPChhd)  
* [Microbiome R package tutorials](http://microbiome.github.io/microbiome/)  

```{r}

sessionInfo()
```


## Package requirements  

* Depends:
+ [R (>= 3.4.1)](https://www.r-project.org/)
+ [phyloseq](https://joey711.github.io/phyloseq/index.html)
+ [ggplot2](http://ggplot2.org/)
+ [microbiome](https://bioconductor.org/packages/devel/bioc/html/microbiome.html)

* Imports:
+ [plyr](https://cran.r-project.org/web/packages/plyr/index.html)
+ [reshape2](https://cran.rstudio.com/web/packages/reshape2/index.html)
+ [stats](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html)
+ [tidyr](http://tidyr.tidyverse.org/)
+ [utils](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/00Index.html)
+ [vegan](https://cran.r-project.org/web/packages/vegan/index.html)
+ [ggpubr](https://cran.r-project.org/web/packages/ggpubr/index.html)
+ [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/index.html)
+ [viridis](https://cran.r-project.org/web/packages/viridis/index.html)
+ [pheatmap](https://cran.r-project.org/web/packages/pheatmap/index.html)
+ [ggrepel](https://cran.r-project.org/web/packages/ggrepel/index.html)

* Suggests:
+ [BiocGenerics](https://bioconductor.org/packages/release/bioc/html/BiocGenerics.html)
+ [knitr](https://cran.r-project.org/web/packages/knitr/index.html)
+ [rmarkdown](http://rmarkdown.rstudio.com/)
+ [ape](https://cran.r-project.org/web/packages/ape/index.html)
+ [Picante](https://academic.oup.com/bioinformatics/article/26/11/1463/203321)
