---
title: "Longitudinal data analysis and visualization"  
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal data analysis and visualization}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```



Longitudinal data presents challenges for visualization.    

## Microbiota plasticity   
### Calculate plasticity  

See Grembi, J.A., Nguyen, L.H., Haggerty, T.D. et al. Gut microbiota plasticity is correlated with sustained weight loss on a low-carb or low-fat dietary intervention. [Sci Rep 10, 1405 (2020).](https://www.nature.com/articles/s41598-020-58000-y)  

Data from `microbiome` package is used here for example.  
```{r, message=FALSE, warning=FALSE, fig.align='center'}
library(microbiome)
library(microbiomeutilities)
library(dplyr)
library(ggpubr)
data(peerj32)
pseq <- peerj32$phyloseq
pseq.rel <- microbiome::transform(pseq, "compositional")
pl <- plasticity(pseq.rel, dist.method = "bray", participant.col="subject")
head(pl)
```

Alternatively, using correlation methods for `plasticity`, one can check for similarity between technical replicates (2X sequenced same sample) for quality check.  

### Plot plasticity  
```{r fig.height=3, fig.width=4, fig.align='center'}

ggplot(pl, aes(group_2,bray)) + 
  geom_boxplot(aes(fill=group_2), 
               alpha=0.5, 
               na.rm = TRUE, 
               width=0.5) +
  geom_jitter(aes(color=group_2), 
              alpha=0.5, 
              size=3, 
              na.rm = TRUE) +
  scale_fill_manual(values = c("#457b9d", "#e63946"))+
  scale_color_manual(values = c("#457b9d", "#e63946"))+
  stat_compare_means() + 
  theme_biome_utils()

```
   
### Area plot    
Area plots can be used to visualize changes in abundances in individual participants, or bioreactors sampled over time.   
Here, we use a randomly chosen small subset of data from [HMP2data](https://bioconductor.org/packages/release/data/experiment/html/HMP2Data.html).  

```{r fig.width=8, fig.height=4, fig.align='center', message=FALSE, warning=FALSE}
library(microbiomeutilities)
library(RColorBrewer)
data("hmp2")
ps <- hmp2
ps.rel <- microbiome::transform(ps, "compositional") 
# chose specific participants to plot data.
pts <- c("Participant_1","Participant_2",
         "Participant_8","Participant_5")
#pts <- "Participant_1"
ps.rel <- subset_samples(ps.rel, subject_id %in% pts)
p <- plot_area(ps.rel, xvar="visit_number", 
               level = "Family",
               facet.by = "subject_id",
               abund.thres = 0.1,
               prev.thres=0.1,
               fill.colors=brewer.pal(12,"Paired"),
               ncol=2,
               nrow=2)
p + ylab("Relative abundance") + 
  scale_y_continuous(labels = scales::percent)
```
 
 
### Paired abundances   
We use a subset of [HMP2 data](https://bioconductor.org/packages/release/data/experiment/html/HMP2Data.html)     
```{r fig.width=5, fig.height=3, message=FALSE, fig.align='center'}
library(microbiome)
library(microbiomeutilities)
#library(gghalves)
library(tidyr)

data("hmp2")
ps <- hmp2 
# pick visit 1 and 2
ps.sub <- subset_samples(ps, visit_number %in% c(1,2))
ps.sub <- prune_taxa(taxa_sums(ps.sub)>0, ps.sub)

ps.rel <- microbiome::transform(ps.sub, "compositional")
ps.rel.f <- format_to_besthit(ps.rel)

# Check how many have both time points
table(meta(ps.rel.f)$visit_number)
#table(meta(ps.rel.f)$visit_number, meta(ps.rel.f)$subject_id)
# there are two Participant_3 and Participant_11 with no time point2 sample 

select.taxa <- c("4430843:Prevotella", "580629:Bacteroides")

group.colors = c("brown3", "steelblue", "grey70")


p <- plot_paired_abundances (ps.rel.f,
                             select.taxa=select.taxa,
                             group="visit_number",
                             group.colors=group.colors,
                             dot.opacity = 0.25,
                             dot.size= 2,
                             add.violin = TRUE,
                             line = "subject_id",
                             line.down = "#7209b7",
                             line.stable = "red",
                             line.up = "#14213d", 
                             line.na.value = "red",
                             line.guide = "none",
                             line.opacity = 0.25,
                             line.size = 1)
print(p + xlab("Time point") + ylab("Relative abundance"))

```

The lines are colored according to their change in abundance from time 1 to time 2.   

### Spaghetti plots   

**One participant many taxa**    
Reference for visualization and original code [ Data to Viz.com](https://www.data-to-viz.com/caveat/spaghetti.html)

The ordering of panel on top of each other is better for comparisons. However, practical consideration can be made on number of columns and rows to distribute the panels.   

```{r message=FALSE,fig.width=4, fig.height=6, message=FALSE, fig.align='center'}
library(microbiomeutilities)
data("hmp2")
pseq <- hmp2 # Ren
pseq.rel <- microbiome::transform(pseq, "compositional")
pseq.relF <- format_to_besthit(pseq.rel)

# Choose one participant
phdf.s <- subset_samples(pseq.relF, subject_id==
                           "Participant_1")
# Choose top 12 taxa to visualize
ntax <- top_taxa(phdf.s, 4)
phdf.s <- prune_taxa(ntax, phdf.s)

plot_spaghetti(phdf.s, plot.var= "by_taxa",
               select.taxa=ntax,
               xvar="visit_number",
               line.bg.color="#8d99ae",
               focus.color="brown3",
               focus.line.size = 1,
               ncol=1,
               nrow=4,
               line.size=0.2)
```

**One taxa many participants**    

```{r message=FALSE,fig.width=6, fig.height=6, message=FALSE, fig.align='center'}

pseq.relF <- format_to_besthit(pseq.rel)
ntax2 <- core_members(pseq.relF, 0.001, 0.5)
# chose first for example
ntax2 # only
# check how many participants are there in "subject_id" 
length(unique(meta(pseq.relF)$subject_id))

# There are 13 participants. Choose ncol and nrow accordingly 

plot_spaghetti(pseq.relF, plot.var= "by_sample",
               select.taxa=ntax2,
               group= "subject_id",
               xvar="visit_number",
               line.bg.color="#8d99ae",
               focus.color="brown3",
               focus.line.size = 1,
               ncol=4,
               nrow=6,
               line.size=0.2)


```

This package is part of the `microbiomeverse` tools.  
See also [microbiome R/BioC package](https://microbiome.github.io/tutorials/)  
**Contributions are welcome:**  

[Issue Tracker](https://github.com/microsud/microbiomeutilities/issues)    
[Pull requests](https://github.com/microsud/microbiomeutilities/pulls)     
[Star us on the Github page](https://github.com/microsud/microbiomeutilities) 


```{r}

sessionInfo()
```